\documentclass[11pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{geometry}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{booktabs}
\usepackage{enumitem}
\usetikzlibrary{calc, arrows.meta, positioning, shapes.geometric, fit, decorations.pathreplacing}

\geometry{margin=2.2cm}

\lstdefinelanguage{Rust}{
  keywords={fn,let,mut,for,in,if,else,while,loop,match,return,struct,impl,pub,use,mod,as,where,move,ref,self,Self,crate,super,async,await,unsafe,const,static,type,enum,trait,dyn,true,false,continue,break},
  keywordstyle=\color{blue!70!black}\bfseries,
  ndkeywords={Vec,Vec3,Mat4,Mesh,MeshVertex,SceneNode,Scene,PbrMaterial,TreeParams,u32,usize,f32,f64,bool,Option,Some,None,Result,Ok,Err,String,Morphology,Tree,k,PI},
  ndkeywordstyle=\color{purple!70!black}\bfseries,
  sensitive=true,
  comment=[l]{//},
  morecomment=[s]{/*}{*/},
  commentstyle=\color{green!50!black}\itshape,
  stringstyle=\color{red!70!black},
  morestring=[b]",
}

\lstset{
  basicstyle=\ttfamily\footnotesize,
  breaklines=true,
  frame=single,
  language=Rust,
  showstringspaces=false,
  tabsize=2,
}

\title{TPMS Shell and Skeletal Morphology Generation\\in engvis}
\author{engvis}
\date{\today}

\begin{document}
\maketitle

\begin{abstract}
Triply Periodic Minimal Surfaces (TPMS) can be used in three distinct
morphology modes for engineering applications: \emph{minimal surface}
(the classic open sheet $f=0$), \emph{shell} (a thick-walled hollow
structure $g = \max(f-\delta,\;-f-\delta) = |f|-\delta$), and
\emph{skeletal} (a solid strut network $f - C = 0$).  This document
presents the mathematical foundation for each mode, the per-surface
thickness calibration constants, the new Split-P surface, the
boundary-based face capping algorithm that produces closed (boundary-free)
meshes, and the implementation pipeline in engvis.
\end{abstract}

% =====================================================================
\section{TPMS Morphology Types}

A triply periodic minimal surface is defined as the zero level-set of
a smooth function $f\colon\mathbb{R}^3 \to \mathbb{R}$:
\begin{equation}\label{eq:tpms-levelset}
  \mathcal{S} = \bigl\{\mathbf{x} \in \mathbb{R}^3 \;\big|\; f(\mathbf{x}) = 0\bigr\}.
\end{equation}
This open surface divides space into two regions ($f>0$ and $f<0$).
Depending on the engineering application, three morphology modes are
useful (see Figure~\ref{fig:morphology-comparison}).

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    every node/.style={font=\small},
    box/.style={draw, minimum width=30mm, minimum height=22mm,
                rounded corners=1mm, align=center},
  ]
  \node[box, fill=blue!8]  (min) {Minimal surface\\$f(\mathbf{x})=0$\\[2pt]\footnotesize open sheet};
  \node[box, fill=orange!10, right=8mm of min] (shell)
    {Shell\\$f^2-\delta^2\le 0$\\[2pt]\footnotesize solid wall};
  \node[box, fill=green!10, right=8mm of shell] (skel)
    {Skeletal\\$f(\mathbf{x})-C=0$\\[2pt]\footnotesize solid strut};

  \draw[->, thick] (min) -- node[above, font=\tiny]{CSG offset} (shell);
  \draw[->, thick] (shell) -- node[above, font=\tiny]{iso-shift} (skel);
\end{tikzpicture}
\caption{Three TPMS morphology modes: minimal surface (open sheet),
shell (two offset surfaces forming a hollow wall), and skeletal
(shifted iso-level forming a solid strut network).}
\label{fig:morphology-comparison}
\end{figure}

\subsection{Minimal Surface Mode}

The default mode extracts an iso-level-set of the TPMS function:
\begin{equation}
  g(\mathbf{x}) = f(\mathbf{x}) - C,
\end{equation}
where $C$ is the iso-level parameter.  The meshed surface is
$\{g = 0\} = \{f = C\}$.  Setting $C = 0$ recovers the classic minimal
surface (the zero level-set); non-zero $C$ shifts the extracted sheet
to a neighbouring iso-surface.

\paragraph{Volume-fraction parameterisation.}\label{sec:vol-frac}
Different TPMS functions have wildly different value ranges
(e.g.\ gyroid $\in[-1.5,1.5]$, neovius $\in[-13,13]$), so a fixed
$C$ slider range cannot cover all surfaces.  engvis therefore exposes
a single \emph{volume fraction} parameter $\varphi\in(0,1)$ with
uniform geometric semantics across all TPMS:
\begin{equation}\label{eq:vol-frac}
  \varphi \;=\; \frac{\bigl|\{\mathbf{x}\in[-1,1]^3 : f(\mathbf{x}) < C\}\bigr|}
                     {|\text{domain}|}.
\end{equation}
The user sets $\varphi$; engvis internally solves for the
corresponding $C$ via \emph{sort-based inversion}: sample $f$ on a
$48^3$ grid, sort the values, and pick the $\varphi$-quantile.  This
makes $\varphi=0.5$ the symmetric minimal surface for \emph{every}
TPMS, and $\varphi\to0$/$\varphi\to1$ the fully filled/void limits.

This produces a thin, open surface---the classic TPMS sheet.  It is
useful for visualisation and mathematical study but not directly for
additive manufacturing, where finite wall thickness is required.

\paragraph{Shared $\varphi$ with skeletal mode.}
The same $\varphi$ parameter drives both the minimal-surface level-set
($f = C$) and the skeletal solid offset (\S\ref{sec:skeletal}).  This
lets the user pick a visually pleasing iso-surface in minimal-surface
mode and then switch directly to skeletal mode to cap it into a closed
solid, without re-tuning the parameter.

\subsection{Shell Mode}\label{sec:shell}

Shell mode produces a \emph{thick-walled solid structure}: a thin
band of material straddling the minimal surface $f=0$, bounded by the
two offset iso-surfaces $f = \pm\delta$.  The solid wall region is the
sub-level set
\begin{equation}\label{eq:shell}
  \mathcal{S}_{\text{shell}}
    = \bigl\{\,\mathbf{x} : f(\mathbf{x})^2 - \delta^2 \le 0 \,\bigr\}
    = \bigl\{\,\mathbf{x} : |f(\mathbf{x})| \le \delta \,\bigr\},
\end{equation}
described by the \textbf{smooth} field $g = f^2 - \delta^2$, which is
negative \emph{inside} the wall and positive outside.

\paragraph{Geometric thickness normalisation.}
The user-facing parameter $t$ is a \emph{geometric} wall thickness
(the physical distance across the wall), not a field offset.  Near the
surface, a TPMS $f(k\mathbf{x})$ has $\|\nabla f\| \approx k$ (the
period factor), so a geometric half-thickness $d$ maps to a field
offset $\delta = d\,\|\nabla f\| \approx d\,k$.  engvis therefore
sets
\begin{equation}\label{eq:thickness}
  \delta = \tfrac{1}{2}\, t \, k,
  \qquad k = \text{tpms\_period},
\end{equation}
so that the meshed wall has true geometric thickness $\approx t$.  For
non-TPMS surfaces, $\|\nabla f\| \approx 1$ and $k$ falls back to $1$.

\paragraph{Why $f^2-\delta^2$, not $|f|-\delta$.}
Both fields share the zero-set $\{|f|=\delta\}$, but $|f|$ has a
$C^1$ cusp at $f=0$: within a single MC~33 cell straddling $f=0$, the
field $|f|-\delta$ has a non-differentiable fold.  The squared field
$f^2-\delta^2$ is smooth everywhere, so MC~33's linear interpolation
stitches adjacent cells without artefacts---\emph{provided the cell
size is smaller than the wall thickness}.

\paragraph{Resolution requirement.}
Because the wall is pierced on both sides within the same region of
space, MC~33 can only reconstruct it when the grid cell is
comfortably smaller than the wall: empirically $\Delta_{\text{cell}}
\lesssim t/5$.  engvis sizes the grid adaptively to
$R \approx 10/t$ (coefficient $10$, floor $96$); if the wall is
thinner than the resolution cap can resolve, the thickness becomes the
limiting feature and the grid is bumped accordingly.

\begin{table}[h]
\centering
\caption{Shell mesh topology for gyroid at geometric thickness
$t=0.1$ ($k=4 \Rightarrow \delta=0.2$).  Once the cell size drops
below the wall thickness ($R\ge 96$), the mesh is a single watertight
closed solid.}
\label{tab:shell-fragments}
\begin{tabular}{lccccc}
\toprule
Field & $R$ & $\chi$ & components & boundary edges & watertight \\
\midrule
$f^2-\delta^2$ & 64  & $-53$ & 9 & 82 & no  \\
$f^2-\delta^2$ & 96  & $-38$ & \textbf{1} & \textbf{0} & \textbf{yes} \\
$f^2-\delta^2$ & 128 & $-38$ & \textbf{1} & \textbf{0} & \textbf{yes} \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Closed solid via box-CSG.}
The wall field is CSG-intersected with the unit-box solid,
$g = \max(f^2-\delta^2,\ \text{box\_sdf})$, before meshing.  Marching
Cubes then generates cap faces on $x/y/z = \pm 1$ itself, yielding a
single watertight closed solid with \emph{zero boundary edges}.  The
Euler characteristic $\chi=-38$ reflects the high genus of the gyroid
wall (its many interconnected channels), as expected for a closed
surface of that topology.

\paragraph{Cap-face boundary smoothing via Newton projection.}
\label{sec:snap-boundary}
The CSG field $g = \max(f,\,\mathrm{sdf}_{\text{box}})$ makes Marching
Cubes generate \emph{cap faces} on each box face $x/y/z = \pm c$.  The
\emph{silhouette} of each cap region --- where the flat cap meets the
curved TPMS surface --- is the intersection curve
\begin{equation}
  C \;=\; \{f = 0\} \,\cap\, \{\text{box face}\},
\end{equation}
a genuine geometric boundary.  MC, however, samples on a fixed 3D
lattice and places these silhouette vertices on \emph{cell-grid
points} rather than on $C$.  Because both free (in-face) coordinates
are locked to the lattice, the field value $f$ at such a vertex is
\emph{not} zero (measured $\mathrm{RMS}|f|\approx 0.043$) and
oscillates from vertex to vertex, producing the visible staircase
along the TPMS/box silhouette (see Fig.~\ref{fig:staircase}).  Note
that this is distinct from the \emph{interior} cap vertices, which sit
on the lattice with $f<0$ and correctly tessellate the flat cap.

\begin{figure}[h]
\centering
\begin{tikzpicture}[scale=1.8]
  % Grid
  \draw[gray!30, very thin] (0,0) grid (4,3);
  % True iso-curve (smooth)
  \draw[blue, thick, smooth] plot coordinates
    {(0.5,0) (1.2,0.5) (2.0,1.0) (2.8,1.5) (3.5,2.0) (4,2.3)};
  % Staircase approximation (MC vertices on cell edges)
  \draw[red, thick] (0.5,0) -- (0.5,1) -- (1.5,1) -- (1.5,2) -- (2.5,2)
    -- (2.5,2.5) -- (3.5,2.5) -- (3.5,3);
  % Labels
  \node[blue, above right] at (4,2.3) {curve $C: f{=}0$};
  \node[red, above left] at (3.5,3) {MC staircase};
  \node[below] at (2,-0.3) {box face (2D subspace)};
\end{tikzpicture}
\caption{Marching Cubes places the cap silhouette vertices on
lattice grid points (red staircase) instead of on the true
intersection curve $C = \{f=0\}\cap\{\text{box face}\}$ (blue).}
\label{fig:staircase}
\end{figure}

engvis eliminates this jaggedness with a post-processing
\textbf{Newton projection} step (\texttt{snap\_cap\_boundary\_to\_curve}):
\begin{enumerate}[leftmargin=2em]
  \item \emph{Identify silhouette vertices}: classify each triangle as
        \emph{cap} (all three vertices on the same box face, same
        sign) or \emph{TPMS}.  A silhouette vertex is one shared
        between a cap triangle and a TPMS triangle.
  \item \emph{Detect box features}: a silhouette vertex may sit not
        only on a single box \emph{face} but on a box \emph{edge}
        (two faces, e.g.\ $x = c \wedge y = c$) or \emph{corner}
        (three faces).  These edges and corners are \emph{hard
        geometric boundaries}: the true TPMS/box silhouette $C$ meets
        the box edge exactly where $\{f=0\}$ crosses that edge, and
        must be respected.  Each vertex is tagged with the set of
        faces it lies on; the count $k\in\{1,2,3\}$ of locked axes
        determines its remaining degrees of freedom.
  \item \emph{Constrained Newton iteration} (dispatch on $k$):
        \begin{itemize}
          \item $k=3$ (\emph{corner}): the vertex is fully pinned ---
                no projection.
          \item $k=2$ (\emph{edge}): two coordinates are fixed at
                $\pm c$, leaving one free axis $t$.  Solve the 1-D
                problem $f(t)=0$ along the edge,
                $t \leftarrow t - f/\partial_t f$, so the vertex lands
                on $\{\text{box edge}\}\cap\{f=0\}$.
          \item $k=1$ (\emph{face}): the box-face normal axis is fixed;
                the remaining two axes $(u,v)$ are free.  Solve the
                2-D problem
                \begin{equation}
                  \begin{pmatrix} u \\ v \end{pmatrix}
                  \leftarrow
                  \begin{pmatrix} u \\ v \end{pmatrix}
                  - \alpha\,\frac{f}{\|\nabla_{\!uv} f\|^2}
                  \begin{pmatrix} \partial_u f \\ \partial_v f \end{pmatrix},
                \end{equation}
                where $\nabla_{\!uv} f = (\partial_u f,\, \partial_v f)$
                is the finite-difference gradient on the free axes.
        \end{itemize}
        In every case the damping factor $\alpha\le 1$ caps each step
        at $\approx 2$ cells, and the free coordinate(s) are
        \textbf{clamped to $[-c, c]$} after each step so the iterate
        cannot slide along $C$ past the face/edge boundary and push
        the mesh outside the box $[-1,1]^3$.
  \item \emph{Accept or reject}: write back the projected position
        only if the total displacement is finite and below
        $\approx 3$ cells; otherwise the iterate is discarded (Newton
        diverged or jumped to another branch of $C$), keeping the
        original lattice vertex inside the box.
\end{enumerate}

This moves each silhouette vertex onto the \emph{exact} curve $C$ ---
including the constrained box-edge intersections --- \emph{without}
changing mesh topology and without ever leaving the box: the measured
mesh bounding box is $[-c,c]^3$ with zero vertices outside, and
$\mathrm{RMS}|f|$ drops sharply.  The
cap triangles remain planar (all vertices keep the locked
coordinate $\pm c$), and the TPMS triangles inherit the smoothed
silhouette.

\paragraph{Vertex splitting and topological watertightness.}
The vertex-splitting step (\S\ref{sec:smart-normals}) intentionally
breaks topological watertightness: boundary vertices are duplicated
so that cap and TPMS triangles no longer share indices.  After
splitting, the mesh has boundary edges along the TPMS/box crease.
This is \emph{by design} --- the mesh remains geometrically closed
(both surfaces face outward), and the split ensures correct shading
at the crease.  The watertightness reported in
Table~\ref{tab:shell-fragments} refers to the pre-split mesh.

\paragraph{Gradient and normals.}
The wall surface $\{f^2=\delta^2\}$ has gradient
$\nabla(f^2-\delta^2) = 2f\,\nabla f$, which is non-degenerate on the
surface itself (where $|f|=\delta\neq 0$).  The outward normal is
$\operatorname{sgn}(f)\,\nabla f / \|\nabla f\|$ (pointing away from
the wall centre $f=0$).  engvis evaluates $f$ at each vertex and
applies this sign correction, producing correct two-sided lighting.

\paragraph{Boundary shading: vertex splitting for the C¹ crease.}
\label{sec:smart-normals}
When the shell field is CSG-intersected with the box solid
($g = \max(f^2-\delta^2,\ \text{box\_sdf})$), the resulting composite
surface has a \emph{C¹-discontinuity} along the TPMS/box boundary
curve: the surface tangent plane jumps from the TPMS tangent to the
flat box face.  This is a \emph{geometric} crease --- the boundary is a
sharp edge, not a smooth blend.

Assigning a single normal per vertex cannot represent this crease:
\begin{itemize}
  \item TPMS gradient $\nabla f$ $\to$ cap faces show spurious curvature
        (not flat)
  \item Face normal $\pm\hat{\mathbf{e}}$ $\to$ TPMS surface shows flat
        patches near the edge
  \item Averaged normal $\to$ both regions look wrong (jagged shading)
\end{itemize}

engvis therefore uses \textbf{vertex splitting} (implemented as
\texttt{overwrite\_smart\_normals}).  Boundary vertices shared between
cap triangles and TPMS triangles are duplicated, and each copy receives
the normal appropriate to its triangle group:

\begin{enumerate}[leftmargin=2em]
  \item \emph{Classify triangles}: a triangle is a ``cap'' triangle if
        all three vertices lie on the same box face (same axis, same
        sign).  Otherwise it is a ``TPMS'' triangle.
  \item \emph{Identify boundary vertices}: a vertex is a boundary
        vertex if it is referenced by both cap and TPMS triangles.
  \item \emph{Split}: for each boundary vertex, create a duplicate.
        The original keeps its index (used by TPMS triangles); the
        duplicate gets a new index (used by cap triangles).
  \item \emph{Remap indices}: update cap triangle indices to point to
        the duplicate vertices.
  \item \emph{Assign normals}:
    \begin{itemize}
      \item TPMS-triangle vertices $\to$ analytic gradient
            $\operatorname{sgn}(f)\,\nabla f / \|\nabla f\|$
            (smooth curved shading; see below)
      \item Cap-triangle vertices $\to$ geometric face normal
            $\pm\hat{\mathbf{e}}$ (flat shading)
    \end{itemize}
\end{enumerate}

\paragraph{Sign correction for shell normals.}
For the shell surface $f^2 - \delta^2 = 0$, i.e.\ $f = \pm\delta$,
the outward normal (pointing away from the wall centre $f=0$) is
$\operatorname{sgn}(f)\,\nabla f$, not $\nabla f$.  On the outer
surface ($f = +\delta$) both agree, but on the inner surface
($f = -\delta$) the raw gradient $\nabla f$ points \emph{inward}
(toward $f=0$), producing incorrect dark shading.  The
\texttt{flip\_inner} flag in \texttt{overwrite\_smart\_normals}
evaluates $f$ at each TPMS vertex and flips the gradient where
$f < 0$, ensuring both sides of the wall are lit correctly.

For skeletal mode ($f = C$), $\nabla f$ already points outward, so
\texttt{flip\_inner} is false.

This produces a clean crease at the boundary --- geometrically correct,
visually sharp, with no jagged artefacts.  The same algorithm is
applied to skeletal mode, where the composite field
$\max(f-C,\ \text{box\_sdf})$ has an identical C¹-discontinuity.

\subsection{Skeletal Mode}\label{sec:skeletal}

Skeletal mode produces a \emph{solid strut network} by shifting the
iso-level of the TPMS function:
\begin{equation}\label{eq:skeletal}
  g(\mathbf{x}) = f(\mathbf{x}) - C,
\end{equation}
where $C$ is the iso-level offset, solved from the user-facing volume
fraction $\varphi$ as described in \S\ref{sec:vol-frac}.

\paragraph{Effect of $\varphi$.}
When $\varphi = 0.5$ (i.e.\ $C$ at the median of $f$), this reduces
to the symmetric minimal surface.  As $\varphi$ increases, the region
$\{g \le 0\} = \{f \le C\}$ expands, filling more of the unit cell
with ``solid'' material.  The resulting surface $f = C$ is
topologically distinct from the minimal surface: instead of an open
sheet, it forms a connected network of solid struts along the edges
and vertices of the TPMS unit cell.

The gradient is simply $\nabla g = \nabla f$, so surface normals are
unaffected by the shift---a desirable property.

\paragraph{Volume-fraction semantics across TPMS.}
Because $C$ is derived from $\varphi$ (not set directly), the same
$\varphi$ value produces comparable solid/void ratios across all
TPMS.  For example, $\varphi=0.3$ always means ``30\% solid, 70\%
void'', regardless of whether the TPMS is gyroid (range $[-1.5,1.5]$)
or neovius (range $[-13,13]$).  This eliminates the need for
per-surface $C$ range tables and makes the UI parameter
semantically uniform.

\paragraph{Skeletal mesh generation with Newton boundary smoothing.}
\label{sec:skeletal-mesh}
The skeletal mesh uses the same CSG-intersect approach as shell mode:
the field $g = \max(f - C,\ \text{box\_sdf})$ is meshed with MC33,
producing a watertight solid with cap faces on $x/y/z = \pm 1$.
However, MC33 places cap-face boundary vertices on discrete grid
edges, causing staircase aliasing along the TPMS/cap junction curve
$C = \{f = C\} \cap \{\text{box face}\}$.

To eliminate this aliasing, engvis applies a \emph{Newton projection}
post-processing step (\texttt{snap\_mesh\_boundary\_to\_curve}) that
moves junction vertices onto the smooth curve $C$:

\begin{enumerate}[leftmargin=2em]
  \item \emph{Identify junction vertices}: for each vertex, check if
        it lies on a box face ($|x_i| \approx 1$ for some axis $i$)
        and is near the isosurface ($|f| < 0.3$).  These are the
        vertices where the TPMS surface meets the cap face.
  \item \emph{Newton projection in 2D free-axis subspace}: for each
        junction vertex, lock the face-normal axis and solve the 2D
        problem
        \begin{equation}
          \begin{pmatrix} u \\ v \end{pmatrix}
          \leftarrow
          \begin{pmatrix} u \\ v \end{pmatrix}
          - \frac{f}{\|\nabla_{\!uv} f\|^2}
          \begin{pmatrix} \partial_u f \\ \partial_v f \end{pmatrix},
        \end{equation}
        where $(u, v)$ are the free axes on the box face.  The
        iteration converges in $\le 24$ steps, moving the vertex onto
        the true curve $C$.
  \item \emph{Smart normals}: after projection, apply the smart
        normals algorithm (\S\ref{sec:smart-normals}) to handle the
        C$^1$ discontinuity at the TPMS/cap junction: gradient normals
        for TPMS vertices, face normals for cap vertices, with vertex
        splitting at the crease.
\end{enumerate}

This approach produces a smooth, watertight skeletal mesh with no
staircase aliasing along the TPMS/cap junction, while preserving the
topological watertightness of the CSG-intersect mesh (boundary edges
$= 0$ before vertex splitting).

\paragraph{Crease snapping for shell and skeletal-solid meshes.}
The same staircase artefact appears wherever a CSG-intersected TPMS
field meets a box cap face, not only in skeletal-network mode.
\texttt{snap\_mesh\_boundary\_to\_curve} is therefore applied in all
three crease-producing paths, with the projection field chosen to
match the crease curve:
\begin{itemize}[leftmargin=2em]
  \item \emph{Shell} (\texttt{build\_shell\_mesh}): the wall surface is
        $f^2 - (t/2)^2 = 0$, so the crease is
        $C = \{f^2 - (t/2)^2 = 0\} \cap \{\text{box face}\}$.  The
        snapping projects against the \emph{wall} field
        $g_{\text{wall}} = f^2 - (t/2)^2$, not $f$ itself.
  \item \emph{Skeletal solid} (\texttt{build\_mesh}, solid branch) and
        the dedicated \texttt{build\_skeletal\_mesh}: the crease is
        $C = \{f = 0\} \cap \{\text{box face}\}$, projected against the
        TPMS field $f$.
\end{itemize}
Because \texttt{snap\_mesh\_boundary\_to\_curve} and
\texttt{project\_to\_curve} are generic over the implicit shape, the
only per-path difference is which compiled field is passed in.  The
Newton iteration and the $\le 3\,h$ displacement guard
($h = \text{cell size}$) reject divergent moves identically in every
case.

% =====================================================================
\section{Thickness Calibration}\label{sec:calibration}

The user-facing thickness $t$ is a true \emph{geometric} wall
thickness.  To realise it consistently across surfaces, engvis
converts $t$ into a field offset $\delta$ using the gradient magnitude
of the implicit function near its zero-set (Eq.~\ref{eq:thickness}):
\begin{equation}
  \delta = \tfrac{1}{2}\, t \, \|\nabla f\|,
  \qquad \|\nabla f\| \approx k = \text{tpms\_period}.
\end{equation}

For a TPMS evaluated as $f(k\mathbf{x})$, the chain rule gives
$\|\nabla_{\mathbf{x}} f\| = k\,\|\nabla_{\mathbf{u}} \hat f\|$, and
the dimensionless gradient $\|\nabla_{\mathbf{u}}\hat f\|$ is $O(1)$
on the surface for all the standard TPMS types
(Table~\ref{tab:calibration}).  Using $k$ as a first-order estimate
therefore yields a wall whose geometric thickness is within a few
percent of $t$---accurate enough for visualisation and consistent
across surfaces without per-surface tuning.

\begin{table}[h]
\centering
\caption{Dimensionless surface-gradient magnitude
$\|\nabla_{\mathbf{u}}\hat f\|$ near the zero-set (the residual
correction absorbed by the $k$ estimate).}
\label{tab:calibration}
\begin{tabular}{lcc}
\toprule
\textbf{Surface} & $\|\nabla_{\mathbf{u}}\hat f\|$ & \textbf{Notes} \\
\midrule
Gyroid           & $\approx 1.2$ & three-term product \\
Schwarz D        & $\approx 1.3$ & four-term sum \\
Schwarz P        & $\approx 1.0$ & simple cosine sum \\
Schoen IWP       & $\approx 1.1$ & \\
Neovius          & $\approx 1.2$ & \\
\bottomrule
\end{tabular}
\end{table}

Because all values are $O(1)$, the single estimate
$\|\nabla f\| \approx k$ suffices; no per-surface calibration table is
required.

% =====================================================================
\section{Split-P Surface}\label{sec:split-p}

The Split-P surface is a TPMS with the symmetry of the Schwarz
P-surface but a more complex topology.  It was included in the
TPMSgen tool but was previously absent from engvis.  Its implicit
equation is:
\begin{equation}\label{eq:split-p}
\begin{aligned}
  f_{\text{Split-P}}(\mathbf{x}) &=
    1.1\bigl[\sin(2kx)\cos(ky)\sin(kz)
           + \sin(kx)\sin(2ky)\cos(kz)
           + \cos(kx)\sin(ky)\sin(2kz)\bigr] \\
  &\quad - 0.2\bigl[\cos(2kx)\cos(2ky)
                   + \cos(2ky)\cos(2kz)
                   + \cos(2kz)\cos(2kx)\bigr] \\
  &\quad - 0.4\bigl[\cos(2kx) + \cos(2ky) + \cos(2kz)\bigr].
\end{aligned}
\end{equation}

Split-P shares structural similarities with the Lidinoid (both contain
$\sin(2x)\cos(y)\sin(z)$ terms) but has distinct second-harmonic
coefficients and an additional linear cosine term.  It belongs to the
same family of surfaces generated by linear combinations of symmetry-adapted
Fourier basis functions.

% =====================================================================
\section{Complete Surface Catalogue}

Table~\ref{tab:catalogue} lists all eleven TPMS surfaces supported by
engvis with their implicit equations $f(\mathbf{x}) = 0$, where
$\mathbf{x} = (kx, ky, kz)$ and $k = 2\pi/\lambda$ is the wave number
controlled by the Period slider.

\begin{table}[h]
\centering
\caption{TPMS surface catalogue in engvis.}
\label{tab:catalogue}
\small
\begin{tabular}{lp{9cm}}
\toprule
\textbf{Name} & \textbf{Implicit equation $f = 0$} \\
\midrule
Gyroid & $\sin x\cos y + \sin y\cos z + \sin z\cos x$ \\
Schwarz P & $\cos x + \cos y + \cos z$ \\
Schwarz D & $\sin x\sin y\sin z + \sin x\cos y\cos z + \cos x\sin y\cos z + \cos x\cos y\sin z$ \\
Schoen IWP & $2(\cos x\cos y + \cos y\cos z + \cos z\cos x) - (\cos 2x + \cos 2y + \cos 2z)$ \\
Neovius & $3(\cos x + \cos y + \cos z) + 4\cos x\cos y\cos z$ \\
F-RD & $4\cos x\cos y\cos z - (\cos 2x\cos 2y + \cos 2y\cos 2z + \cos 2z\cos 2x)$ \\
Lidinoid & $\tfrac{1}{2}[\sin 2x\cos y\sin z + \sin x\sin 2y\cos z + \cos x\sin y\sin 2z]
            - \tfrac{1}{2}[\cos 2x\cos 2y + \cos 2y\cos 2z + \cos 2z\cos 2x] + 0.15$ \\
Split-P & see Eq.~\eqref{eq:split-p} \\
Fischer-Koch S & $\cos 2x\sin y\cos z + \cos 2y\sin z\cos x + \cos 2z\sin x\cos y$ \\
Fischer-Koch Y & $2\cos x\cos y\cos z + \sin 2x\sin y + \sin 2y\sin z + \sin 2z\sin x$ \\
Fischer-Koch CP & $\cos x + \cos y + \cos z + 4\cos x\cos y\cos z$ \\
\bottomrule
\end{tabular}
\end{table}

% =====================================================================
\section{Implementation Pipeline}

The morphology transformation is applied as a post-processing step on
the base TPMS tree, before meshing.  Figure~\ref{fig:pipeline} shows
the complete data flow.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    every node/.style={font=\small},
    block/.style={draw, rounded corners=1.5mm, minimum height=8mm,
                  align=center, fill=blue!6},
    morph/.style={draw, rounded corners=1.5mm, minimum height=8mm,
                  align=center, fill=orange!10},
    mesh/.style={draw, rounded corners=1.5mm, minimum height=8mm,
                 align=center, fill=green!6},
    arr/.style={->, thick, >=Stealth},
  ]

  \node[block] (tree) {build\_tree(params)\\$\downarrow$\\base $f(\mathbf{x})$};
  \node[morph, below=8mm of tree] (morph) {Morphology transform\\$\downarrow$\\$g(\mathbf{x})$};
  \node[mesh, below=8mm of morph] (mc) {MC33 / DC meshing\\$\downarrow$\\Mesh};
  \node[block, below=8mm of mc] (scene) {Scene + Render};

  \draw[arr] (tree) -- (morph);
  \draw[arr] (morph) -- (mc);
  \draw[arr] (mc) -- (scene);

  % Side annotations
  \node[right=12mm of morph, font=\scriptsize, text=gray, align=left]
    {Minimal: $g = f$\\Shell: $g = f^2-\delta^2$\\Skeletal: $g = f - C$};
\end{tikzpicture}
\caption{Data flow for TPMS morphology generation.  The base TPMS
function $f$ is first constructed by \texttt{build\_tree}, then
transformed by the morphology module into $g$, and finally meshed by
the standard MC33 or DC pipeline.}
\label{fig:pipeline}
\end{figure}

\subsection{Code Structure}

The morphology transformation is implemented in \texttt{build\_scene}.
For minimal surface and skeletal modes, the tree is transformed
algebraically and passed to the standard \texttt{build\_mesh}
pipeline.  For shell mode, the original tree is passed to
\texttt{build\_shell\_mesh}, which meshes the smooth wall solid
$f^2-\delta^2$:

\begin{lstlisting}
// build_scene: geometric thickness -> field offset
let k = self.tpms_period.max(1.0);     // |grad f| ~ k
let shell_half_t = 0.5 * thickness * k; // delta

let tree = match self.morphology {
    MinimalSurface => tree - iso_level, // f = C level-set
    Shell => tree,                       // handled below
    Skeletal => tree - iso_level,       // f = C solid
};


let mesh = if matches!(self.morphology, Morphology::Shell) {
    // Smooth wall solid f^2 - delta^2, box-CSG, single MC33 pass.
    build_shell_mesh(
        tree.clone(), shell_half_t, &label, effective_res,
        clip_to_unit_ball, clip_radius,
    )
} else {
    build_mesh(tree.clone(), &label, ...)
};
\end{lstlisting}

\begin{lstlisting}
// build_shell_mesh: smooth wall solid + box-CSG, single pass
fn build_shell_mesh(tree, half_t, name, res, ...) -> Mesh {
    // Box-CSG: intersect with unit-box solid so MC caps the faces.
    let box_sdf = |x|.max(|y|).max(|z|) - c;

    // Thin-wall solid: g = f^2 - half_t^2 (smooth, no cusp).
    let wall  = tree.square() - half_t * half_t;
    let field = wall.max(box_sdf);
    build_mc33_mesh_domain(field, name, res, false, half)
}
\end{lstlisting}

The Fidget implicit modelling library represents the expression as a
computational tree (\texttt{Tree}) with overloaded arithmetic
operators: \texttt{tree.square()} builds $f^2$, and \texttt{.max()}
performs the CSG intersection with the box solid.  The resulting
smooth field is meshed by the standard MC33 pipeline in a single
pass, yielding a watertight closed wall.

\subsection{Interaction with Meshing}

Both meshing backends (MC33 and Dual Contouring) handle the
morphology-transformed function identically to the base function:
they sample $g(\mathbf{x})$ on a regular grid (MC33) or adaptive
octree (DC) and extract the zero level-set.

For \textbf{shell mode}, the smooth field $g = f^2-\delta^2$ is
CSG-intersected with the unit-box solid
($\max(g,\ \text{box\_sdf})$), so MC33 generates cap faces on
$x/y/z = \pm 1$, producing a closed solid with zero boundary edges.
Because $f^2$ is smooth (no $C^1$ cusp), MC33's linear interpolation
stitches adjacent cells cleanly once the cell size is below the wall
thickness ($\Delta_{\text{cell}} \lesssim t/5$); below that the wall
fragments (Table~\ref{tab:shell-fragments}, $R=64$).  The wall
surface gradient $\nabla(f^2-\delta^2) = 2f\,\nabla f$ is
non-degenerate on the surface (where $|f|=\delta$), giving clean
vertex normals.

For \textbf{skeletal mode}, the function $g = f - C$ has the same
gradient structure as the base TPMS ($\nabla g = \nabla f$), so both
backends produce clean results.  The skeletal mesh is a closed solid
(box-CSG), so it uses the boundary capping described in
Section~\ref{sec:capping}.

\subsection{Async Mesh Building and Performance}\label{sec:async}

High-resolution TPMS meshes (e.g.\ gyroid shell at $R=256$) can produce
millions of triangles, and the MC33 extraction, vertex deduplication,
and topology analysis can take several seconds.  To keep the UI
responsive, engvis uses a \emph{background thread} for mesh building:

\begin{enumerate}[leftmargin=2em]
  \item When a parameter changes (\texttt{needs\_remesh = true}), the
        UI thread creates an \texttt{AppBuildSnapshot} --- a clone of
        all parameters needed for the build.
  \item A background \texttt{thread::spawn} calls
        \texttt{build\_scene\_result()} on the snapshot, producing a
        \texttt{MeshBuildResult} (scene + topology + status).
  \item The result is stored in an \texttt{Arc<Mutex<Option<...>>>}
        shared between the UI thread and the build thread.
  \item Each frame, the UI thread checks whether the result is ready
        (\texttt{Some}).  If so, it swaps the scene and updates the
        status bar.  While building, the status bar shows
        ``\textcolor{orange}{Building mesh...}''.
\end{enumerate}

This approach avoids blocking the egui event loop, preventing the
``not responding'' state that occurs when mesh building runs
synchronously on the UI thread.

\paragraph{Winding-fix optimisation.}
The \texttt{fix\_winding} function's per-component volume flip
previously iterated all faces for each negative-volume component
($O(n_{\text{faces}} \times n_{\text{components}})$).  This is now
$O(n_{\text{faces}})$ by building a root$\to$face-index list
(\texttt{comp\_faces}) during the volume accumulation pass and
iterating only over faces in negative-volume components.

\paragraph{MC33 winding redundancy.}
MC33's \texttt{extract\_par} already applies a gradient-based winding
fix during extraction.  The previous call to \texttt{from\_triangles}
(which runs BFS + union-find + geometric-flip detection) was
redundant for MC33 meshes and has been replaced with
\texttt{from\_triangles\_open}, which only performs vertex
deduplication.  This eliminates the $O(n^2)$ geometric-flip detection
pass on high-resolution meshes.

\paragraph{Edge index deduplication.}
The \texttt{extract\_edge\_indices} function previously stored each
triangle's 3 edges independently, duplicating shared edges on closed
manifolds (2$\times$ overhead).  It now uses a
\texttt{HashSet<(u32, u32)>} to store each edge only once, halving
the GPU edge endpoint buffer size for closed surfaces.

\paragraph{Batch grid evaluation.}
The previous implementation sampled the scalar field by calling
\texttt{f(x, y, z)} per grid point, where each call created a new
fidget VM evaluator.  For a resolution of $R = 256$, this meant
$(257)^3 \approx 17\text{M}$ evaluator creations.

The optimised path (\texttt{extract\_par\_with\_grid}) pre-computes
the entire grid in a single \texttt{float\_slice\_eval} call:
\begin{enumerate}[leftmargin=2em]
  \item Build three coordinate arrays $\mathbf{x}, \mathbf{y},
        \mathbf{z}$ of length $(R+1)^3$.
  \item Call \texttt{eval.eval(\&tape, \&xs, \&ys, \&zs)} once,
        returning all field values in a single batch.
  \item Pass the pre-computed grid to \texttt{extract\_par\_with\_grid},
        which skips internal sampling and proceeds directly to
        per-cell triangulation.
\end{enumerate}

This replaces $O(R^3)$ individual evaluator creations with a single
batch evaluation, eliminating the dominant overhead for
high-resolution meshes.

\paragraph{Parallel winding fix.}
The gradient-based winding fix (comparing each triangle's face normal
against $\nabla f$ at the centroid) was previously sequential.
\texttt{extract\_par\_with\_grid} now parallelises this with
\texttt{rayon::par\_chunks\_exact\_mut}, distributing the 6 field
evaluations per triangle across all CPU cores.

\subsection{Box-Face Capping via Boundary Loops}\label{sec:capping}

Shell and skeletal modes define a \emph{solid region}
($g \le 0$) bounded by the iso-surface and the domain boundary.
The MC33 meshing backend extracts only the iso-surface $\{g = 0\}$,
leaving the mesh open where it meets the six faces of $[-1,1]^3$.
To produce a closed volume, engvis identifies the \emph{MC33 boundary
edges} on the box surface and fan-triangulates the enclosed loops
directly---reusing the existing boundary vertices so that cap and body
share the same vertex indices (Figure~\ref{fig:capping}).

\emph{Note:} Shell mode uses the same box-CSG capping mechanism
(via \texttt{build\_mc33\_mesh\_domain}), applied independently to
each of the two iso-surfaces.  The capping algorithm below describes
the skeletal case; the shell case is identical but applied twice.

\begin{figure}[h]
\centering
\begin{tikzpicture}[
    every node/.style={font=\small},
    arr/.style={->, thick, >=Stealth},
  ]
  % Box face with contour
  \draw[thick] (0,0) rectangle (3.2,3.2);

  % g=0 contour (closed loop)
  \draw[blue, thick] (0.6,0.4) .. controls (0.2,1.5) and (0.3,2.5) .. (1.2,2.9)
    .. controls (2.0,3.1) and (2.8,2.3) .. (2.7,1.4)
    .. controls (2.5,0.5) and (1.5,0.2) .. (0.6,0.4);

  % Centroid and fan lines
  \coordinate (C) at (1.5,1.6);
  \fill[orange!25] (0.6,0.4) .. controls (0.2,1.5) and (0.3,2.5) .. (1.2,2.9)
    .. controls (2.0,3.1) and (2.8,2.3) .. (2.7,1.4)
    .. controls (2.5,0.5) and (1.5,0.2) .. (0.6,0.4);
  \fill[red!60!black] (C) circle (1.5pt);
  \draw[gray!60] (C) -- (0.6,0.4);
  \draw[gray!60] (C) -- (1.2,2.9);
  \draw[gray!60] (C) -- (2.7,1.4);

  % Labels
  \node at (1.6,-0.3) {box surface $[-1,1]^3$};
  \node[blue] at (3.0,2.6) {boundary loop};
  \node[red!60!black] at (1.8,1.8) {\tiny centroid};
\end{tikzpicture}
\caption{Boundary-based face capping: the MC33 boundary edges
(blue) on the box surface are chained into closed loops, then
fan-triangulated from their centroid (red dot) to produce cap
triangles (orange) that close the volume.  Since the cap reuses
the existing MC33 boundary vertex indices, no vertex deduplication
is needed to merge cap and body.}
\label{fig:capping}
\end{figure}

\paragraph{Prerequisites.}
The algorithm requires that all MC33 boundary edges lie on the
surface of $[-1,1]^3$ (no \emph{interior} boundary edges).  For the
CSG shell formula $g = |f| - \delta$, this holds at MC33 resolution
$\ge 96$.  At resolution 64, the old squared formula produced 19
spurious micro-components with $\sim$1500 interior boundary edges
that could not be capped.  The adaptive resolution logic enforces a
minimum of 96 for shell and skeletal modes.

\paragraph{Algorithm.}
\begin{enumerate}[leftmargin=2em]
  \item \textbf{Identify boundary edges}: an edge shared by exactly
    one triangle.  Classify each as \emph{surface} (both endpoints on
    the box surface, $|x_i| > 1 - \varepsilon$) or \emph{interior}.
    If any interior boundary edges exist, abort capping.
  \item \textbf{Build adjacency graph}: vertex $\to$ list of
    neighbouring boundary vertices.  Every vertex must have degree
    exactly 2 (prerequisite for closed simple loops).
  \item \textbf{Chain into closed loops}: greedy traversal from an
    arbitrary unvisited vertex, following unvisited neighbours until
    the loop closes.  Loops may cross box edges (where two cube faces
    meet), so all boundary edges are processed as one global graph
    rather than per-face.
  \item \textbf{Fan-triangulate each loop}: add a centroid vertex
    $\mathbf{c} = \frac{1}{n}\sum \mathbf{v}_i$ and create triangles
    $(\mathbf{c}, \mathbf{v}_i, \mathbf{v}_{i+1})$.  The loop normal
    is estimated via Newell's method:
    \begin{equation}
      \mathbf{n} = \sum_{i=0}^{n-1} (\mathbf{v}_i - \mathbf{v}_{i+1})
                   \times (\mathbf{v}_i + \mathbf{v}_{i+1}),
    \end{equation}
    which handles non-planar loops.  The winding direction is chosen
    so the cap triangle cross products align with $\mathbf{n}$.
  \item \textbf{Signed-volume check}: compute
    $V = \sum_{\triangle} \mathbf{p}_0 \cdot (\mathbf{p}_1 \times \mathbf{p}_2)$.
    If $V < 0$, all triangle indices are swapped (reversing the
    orientation globally).
\end{enumerate}

\paragraph{Why reusing MC33 vertices matters.}
An earlier approach extracted the $g = 0$ contour on each face via
independent Marching Squares (MS), then relied on vertex
deduplication to merge the MS contour vertices with the MC33 boundary
vertices.  Although both methods interpolate the same function on the
same grid points, the separate Shape tape compilation could introduce
sub-ULP differences, causing $\sim$50\% of cap vertices to remain
unmerged and leaving thousands of boundary edges.

The direct boundary approach eliminates this entirely: the cap
triangles reference the \emph{same vertex indices} as the MC33 body,
so no merging is needed.  The only new vertices are the loop
centroids.

\paragraph{Topology verification.}
After capping, the mesh topology is verified via the Euler
characteristic $\chi = V - E + F$ and the boundary edge count.
For the Gyroid shell (period $k=4$, thickness $t=1.0$):
\begin{itemize}[leftmargin=2em]
  \item \emph{Before capping}: $\chi = -20$, 4 components,
    5232 boundary edges (all on box faces), 0 interior boundaries.
  \item \emph{After capping}: $\chi = 0$, 4 components,
    \textbf{0 boundary edges}, 0 non-manifold edges.
\end{itemize}
The 4 components correspond to the two offset surfaces ($f = \pm\delta$)
of the two interpenetrating Gyroid channel networks---a topologically
correct result.  Each component is individually closed
(boundary-free), though the mesh as a whole has $>1$ connected
component.

% =====================================================================
\section{Comparison with TPMSgen}

The design of engvis's morphology system follows the principles
established by TPMSgen (Albert Fores~G., GitHub), adapted for
engvis's Rust/Fidget meshing pipeline.  Key differences are summarised
in Table~\ref{tab:comparison}.

\begin{table}[h]
\centering
\caption{Comparison of TPMSgen and engvis TPMS generation.}
\label{tab:comparison}
\begin{tabular}{lp{4.5cm}p{4.5cm}}
\toprule
\textbf{Feature} & \textbf{TPMSgen} & \textbf{engvis} \\
\midrule
Shell generation & Two MC passes at $\pm\delta$, concat meshes & CSG $\max(f-\delta,\,-f-\delta)$ single pass \\
Skeletal generation & MC at $f=0$, boolean diff with box & MC at $f=C$, box clip \\
Boolean operations & Blender engine (trimesh) & CSG $\max/\min$ on expression tree \\
Meshing backend & skimage marching\_cubes & Custom MC33 + Fidget DC \\
Watertight closure & Iterative padding + hole filling & Boundary-loop fan capping \\
Surface count & 10 (5 shell + 5 skeletal) & 11 (all in all modes) \\
Normal handling & pyvista flip normals & Gradient overwrite + signed-volume check \\
\bottomrule
\end{tabular}
\end{table}

The key architectural difference is that TPMSgen performs explicit
boolean mesh operations (intersection/difference with bounding boxes)
using Blender as a backend, while engvis relies on the implicit
surface being naturally clipped to the $[-1,1]^3$ domain.  This
produces open boundary edges at the domain faces but avoids the
dependency on an external geometry kernel.

% =====================================================================
\section{Usage}

In the engvis UI, morphology controls appear in the \textbf{Source}
panel when a TPMS surface is selected:

\begin{enumerate}[leftmargin=2em]
  \item Select a TPMS surface from the dropdown (e.g.\ Gyroid).
  \item Adjust the \textbf{Period k} slider to set the spatial frequency.
  \item Choose a morphology mode:
    \begin{itemize}
      \item \textbf{Minimal surface} ($f=C$) --- open sheet.
      \item \textbf{Shell} ($|f|-\delta=0$) --- adjust \textbf{Thickness}.
      \item \textbf{Skeletal} ($f-C=0$) --- adjust \textbf{Volume fraction $\varphi$}.
    \end{itemize}
  \item For minimal-surface and skeletal modes, adjust the
        \textbf{Volume fraction $\varphi$} slider ($0.01$--$0.99$).
        $\varphi=0.5$ gives the symmetric minimal surface; lower/higher
        values shift the iso-level to fill or empty the unit cell.
  \item Click \textbf{Re-mesh} to rebuild the surface.
\end{enumerate}

The status bar at the bottom displays the current mode alongside
topology statistics (vertex/edge/face counts, Euler characteristic,
boundary edge count, watertightness).

% =====================================================================
\section{Performance Optimisation}\label{sec:perf}

The mesh generation pipeline has been optimised to achieve interactive
frame rates.  Table~\ref{tab:perf} shows the timing breakdown for a
gyroid at resolution $R=96$ (shell: $121\,639$ vertices, $243\,328$
triangles; skeletal: $36\,100$ vertices, $72\,200$ triangles).

\begin{table}[h]
\centering
\caption{Pipeline timing (release build, Apple M-series, $R=96$).
All times in milliseconds.}
\label{tab:perf}
\begin{tabular}{lcc}
\toprule
Stage & Shell & Skeletal \\
\midrule
Shape (JIT compile)   & 0    & 0  \\
Grid build            & 2    & 0  \\
Grid eval (JIT)       & 14   & 4  \\
MC33 extract          & 3    & 1  \\
Vertex dedup          & 6    & 2  \\
Winding fix           & 28   & 6  \\
Gradient normals      & 5    & 2  \\
\midrule
\textbf{Total}        & \textbf{58} & \textbf{15} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{JIT Evaluation}\label{sec:jit}

The Fidget implicit-function evaluator was migrated from the
stack-based VM (\texttt{VmFunction}) to the JIT compiler
(\texttt{JitFunction}).  For a $97^3$ grid ($912\,673$ points), this
reduced evaluation time from $452$\,ms to $14$\,ms --- a
$\mathbf{32\times}$ speedup.  The JIT compiles the expression tree to
native machine code once, then evaluates all grid points in a single
batched \texttt{float\_slice\_eval} call.

\subsection{Sort-Based Vertex Deduplication}\label{sec:dedup}

Vertex deduplication merges coincident vertices at octree resolution
boundaries.  The previous \texttt{HashMap}-based approach allocated a
bucket per vertex and probed neighbours, costing $\sim 113$\,ms for
$121\,639$ vertices.

The optimised algorithm (\texttt{dedup\_vertices} in
\texttt{mesh.rs}) uses a \emph{sort-and-group} strategy:
\begin{enumerate}[leftmargin=2em]
  \item \emph{Parallel key generation}: for each vertex, compute a
        spatial hash key by quantising coordinates to the dedup
        tolerance $\varepsilon$ and packing into a \texttt{u64}:
        \begin{equation}
          h(\mathbf{p}) = \bigl(\lfloor x/\varepsilon \rceil + 2^{20}\bigr)
                        \,\big|\,\bigl(\lfloor y/\varepsilon \rceil + 2^{20}\bigr) \ll 21
                        \,\big|\,\bigl(\lfloor z/\varepsilon \rceil + 2^{20}\bigr) \ll 42.
        \end{equation}
  \item \emph{Parallel sort}: sort the $(h, \text{idx})$ pairs by
        key using Rayon's \texttt{par\_sort\_unstable\_by\_key}.
  \item \emph{Sequential grouping}: scan the sorted array; vertices
        with the same key are candidates for merging.  For each
        candidate, verify the actual Euclidean distance
        $d^2 < \varepsilon^2$ (to avoid hash-collision false positives)
        before merging.
  \item \emph{Parallel index remap}: rewrite all triangle indices
        using the remap table, in parallel.
  \item \emph{Parallel degenerate removal}: flag triangles with
        repeated indices ($a=b \vee b=c \vee a=c$) for removal, in
        parallel; compact the index buffer sequentially.
\end{enumerate}
This reduced dedup time from $113$\,ms to $6$\,ms --- a
$\mathbf{19\times}$ speedup --- by eliminating per-vertex
\texttt{HashMap} allocations and improving cache locality through
sorted access.

\subsection{Parity Union-Find for Winding Correction}\label{sec:parity-uf}

Winding correction ensures all triangles face outward consistently.
The previous approach used BFS traversal across face adjacency
(separate from a Union-Find for component counting), costing
$\sim 235$\,ms for $243\,328$ triangles.

The optimised algorithm (\texttt{fix\_winding} in \texttt{mesh.rs})
uses a single \emph{Parity Union-Find} that simultaneously computes
winding consistency and connected components:
\begin{enumerate}[leftmargin=2em]
  \item \emph{Edge adjacency}: build a hash map from edge key
        $(\min(a,b), \max(a,b)) \to [(face, direction)]$ using
        \texttt{FxHashMap} for fast hashing.
  \item \emph{Parity Union-Find}: each face has a parity bit (0/1)
        relative to its root.  When two faces $f_a, f_b$ share an edge,
        compute the \emph{edge parity} $p_e = (dir_a == dir_b)$ (1 if
        inconsistent).  Union the two components, setting the link
        parity so that
        \begin{equation}
          parity(f_a \to f_b) = p_a \oplus parity[r_b] \oplus p_b = p_e.
        \end{equation}
  \item \emph{Path compression}: the \texttt{find} function performs
        two-pass path compression, rewiring each node to point directly
        to the root and updating its parity accordingly.
  \item \emph{Flip}: after all unions, flip faces with parity 1
        relative to their root.
  \item \emph{Signed volume}: compute per-component signed volume
        (parallel via Rayon \texttt{fold/reduce}); flip components
        with negative volume.
\end{enumerate}
This reduced winding-fix time from $235$\,ms to $28$\,ms --- a
$\mathbf{8.4\times}$ speedup --- by eliminating the separate BFS pass
and leveraging path compression for near-constant-time amortised
\texttt{find} operations.

\subsection{Topology Computation}\label{sec:topo}

The \texttt{compute\_topology} function was optimised by migrating
from \texttt{std::HashMap<(u32,u32), Vec<usize>} to
\texttt{FxHashMap<u64, Vec<usize>}, using a packed \texttt{u64} key
for faster hashing.  Topology computation is now optional
(enabled by the \texttt{ENGVIS\_TOPO} environment variable) and
skipped in normal interactive use, saving $\sim 22$\,ms per remesh.

\subsection{Selftest Mode}\label{sec:selftest}

The \texttt{-{}-selftest} CLI flag runs an automated test suite that
validates:
\begin{itemize}
  \item \textbf{TPMS value ranges}: samples each TPMS on a $64^3$
        grid and reports $[\min, \max]$ and $\operatorname{vol\_frac}(C=0)$.
  \item \textbf{Volume-fraction solver}: for $\varphi \in \{0.3, 0.5, 0.7\}$,
        solves $C$ and verifies the meshed iso-surface has
        $\operatorname{mean}|f-C| < 0.001$.
  \item \textbf{Skeletal topology}: $V=36\,100$, $F=72\,200$,
        $\chi=0$, boundary edges $=0$.
  \item \textbf{Shell topology}: $V=121\,639$, $F=243\,328$,
        $\chi=-51$, boundary edges $=52$ (pre-split).
  \item \textbf{Timing}: per-stage millisecond breakdown for
        grid build, eval, extract, dedup, winding fix, normals.
\end{itemize}

\end{document}
